Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data

### 2022-12-19 manuscript.tex
Updates from Overleaf

% formatting links in the PDF file
\hypersetup{
pdfpagemode={UseOutlines},
bookmarksopen=true,
bookmarksopenlevel=0,
hypertexnames=false,
colorlinks   = true, %Colours links instead of ugly boxes
urlcolor     = blue, %Colour for external hyperlinks
linkcolor    = blue, %Colour of internal links
citecolor   = cyan, %Colour of citations
pdfstartview={FitV},
unicode,
breaklinks=true,
}

@@

\section{Features of Psychological Time Series}

@@

@@

\subsection{Input variables}
Time series clustering starts with the selection and preparation of the variables of interest. While the selection will necessarily be field- and concept-specific, there are a few conceptual and methodological issues that should be considered. Conceptually, the included variables should adequately capture the concept of interest and should be meaningful to the understanding of the time series. There are, for example, calls that emotion dynamics should be assessed with a repertoire of positive and negative emotions \citep[e.g.,][]{dejonckheere2019}, migration experiences are fully captured with affect, behavior, cognition, and desire measurements \citep[e.g.,][]{Kreienkamp2022d}, and many health developments are commonly captured within the biopsychosocial domains \citep[e.g.,][]{suls2004}. At the same time, however, the added number of variables can become a methodological concern. Not only can redundant and irrelevant variables diminish the quality of the analyses, but with intensive longitudinal data the number of data points compounds across participants, measurement occasions, and variables so that additional variables can make many of the following steps substantially more difficult (also see \fgrref{fig:TSCFlowN}).

@@

\subsection{Feature extraction}
The main aim of feature extraction is to describe the most important and meaningful aspects of a time series. There are, however, hundreds of possible features to describe psychological time series \citep[e.g., tsfresh][]{christ2018}. One approach to choosing relevant features would be to extract a large number of features and then assess which features are most effective at capturing differences in the time series. However, such an approach is not always advisable for psychological time series. For one, features should reduce the data dimensionality --- it would thus not necessarily be advisable to describe 60 ESM measurements with 180 time series features. More importantly, however, careful feature extraction can be crucial in the interpretability and explainability of the results. This is particularly the case when features have a conceptual or psychological meaning. Taking the concept of well-being as an example, psychologists might be interested in whether certain participants tend to have higher well-being in general (i.e., mean) or whether some participants fluctuate between extremely high and low well-being (i.e., variance). But psychologists might not necessarily be interested in the exact time point after which 50\% of the summed well-being values lie (i.e., relative mass quantile index) or how much different sine wave patterns within the well-being data correlate with one another (i.e., cross power spectral density). We would, thus, strongly advocate for a careful selection of time series features that are meaningful to the field and concepts.

@@

\subsection{Feature reduction}
Once a meaningful selection of time series features has been extracted for each variable and participant, the total number of data points usually remains too large for most clustering algorithms. As an example, a relatively common scenario would include 10 variables of interest, where 8 time series features are extracted, resulting in 80 features per participant (with a common sample size of 100 participants that would result in a total of 8,000 data points in this hypothetical example). We offer an illustration of the compounding numbers of data points in \fgrref{fig:TSCFlowN}. The difficulty of working with such a large number of dimensions is sometimes termed the `dimensionality curse' \citep[e.g.,][]{altman2018}. To deal with this dimensionality issue, two main approaches have been proposed --- feature selection and feature projection (a full overview of the approaches, methods, and common algorithms is available in \tblref{tab:featureReduction}).

@@

\subsection{Feature clustering}
For feature-based time-series clustering, the main aim is to organize participants into groups so that the features of participants within a group are as similar as possible, while the features of people in different groups are as different as possible. The crux of clustering is, thus, to have clearly defined and effective measurements of (dis)similarity. Most of the clustering algorithms used today use some form of distance measurement to optimize group assignment (or similarity measurement for qualitative features). While others have produced excellent overviews of the many clustering approaches available \citep[e.g.,][]{xu2015}, we will briefly introduce some of the more readily available approaches suitable for most time series feature data. The well-established and readily available clustering approaches can, broadly speaking, be categorized as based on (1) centroids, (2) distributions, (3) density, (4) hierarchies, or (5) a combination thereof (see \tblref{tab:clusterApproaches}).

@@

\subsection{Cluster Performance Evaluation}

@@

>>>>

% formatting links in the PDF file
\hypersetup{
pdfpagemode={UseOutlines},
bookmarksopen=true,
bookmarksopenlevel=0,
hypertexnames=false,
colorlinks   = true, %Colours links instead of ugly boxes
urlcolor     = blue, %blue,Colour for external hyperlinks
linkcolor    = black, %blue, Colour of internal links
citecolor   = black, % cyan, Colour of citations
pdfstartview={FitV},
unicode,
breaklinks=true,
}

@@

@@

\subsection{Features of Psychological Time Series}

\paragraph{Central Tendency}

\paragraph{Variability}

\paragraph{(In)stability}

\paragraph{Inertia and Periodicity}

\paragraph{Linear Trend}

\paragraph{Nonlinearity}

\paragraph{Analysis Steps}

@@

\paragraph{Input variables}
Time series clustering starts with the selection and preparation of the variables of interest. While the selection will necessarily be field- and concept-specific, there are a few conceptual and methodological issues that should be considered. Conceptually, the included variables should adequately capture the concept of interest and should be meaningful to the understanding of the time series. There are, for example, calls that emotion dynamics should be assessed with a repertoire of positive and negative emotions \citep[e.g.,][]{dejonckheere2019}, migration experiences are fully captured with affect, behavior, cognition, and desire measurements \citep[e.g.,][]{Kreienkamp2022d}, and many health developments are commonly captured within the biopsychosocial domains \citep[e.g.,][]{suls2004}. At the same time, however, the added number of variables can become a methodological concern. Not only can redundant and irrelevant variables diminish the quality of the analyses, but with intensive longitudinal data the number of data points compounds across participants, measurement occasions, and variables so that additional variables can make many of the following steps substantially more difficult (also see \fgrref{fig:TSCFlowN}).

@@

\paragraph{Feature extraction}
The main aim of feature extraction is to describe the most important and meaningful aspects of a time series. There are, however, hundreds of possible features to describe psychological time series \citep[e.g., tsfresh][]{christ2018}. One approach to choosing relevant features would be to extract a large number of features and then assess which features are most effective at capturing differences in the time series. However, such an approach is not always advisable for psychological time series. For one, features should reduce the data dimensionality --- it would thus not necessarily be advisable to describe 60 ESM measurements with 180 time series features. More importantly, however, careful feature extraction can be crucial in the interpretability and explainability of the results. This is particularly the case when features have a conceptual or psychological meaning. Taking the concept of well-being as an example, psychologists might be interested in whether certain participants tend to have higher well-being in general (i.e., mean) or whether some participants fluctuate between extremely high and low well-being (i.e., variance). But psychologists might not necessarily be interested in the exact time point after which 50\% of the summed well-being values lie (i.e., relative mass quantile index) or how much different sine wave patterns within the well-being data correlate with one another (i.e., cross power spectral density). We would, thus, strongly advocate for a careful selection of time series features that are meaningful to the field and concepts.

@@

\paragraph{Feature reduction}
Once a meaningful selection of time series features has been extracted for each variable and participant, the total number of data points usually remains too large for most clustering algorithms. As an example, a relatively common scenario would include 10 variables of interest, where 8 time series features are extracted, resulting in 80 features per participant (with a common sample size of 100 participants that would result in a total of 8,000 data points in this hypothetical example). We offer an illustration of the compounding numbers of data points in \fgrref{fig:TSCFlowN}. The difficulty of working with such a large number of dimensions is sometimes termed the `dimensionality curse' \citep[e.g.,][]{altman2018}. To deal with this dimensionality issue, two main approaches have been proposed --- feature selection and feature projection (a full overview of the approaches, methods, and common algorithms is available in \tblref{tab:featureReduction}).

@@

\paragraph{Feature clustering}
For feature-based time-series clustering, the main aim is to organize participants into groups so that the features of participants within a group are as similar as possible, while the features of people in different groups are as different as possible. The crux of clustering is, thus, to have clearly defined and effective measurements of (dis)similarity. Most of the clustering algorithms used today use some form of distance measurement to optimize group assignment (or similarity measurement for qualitative features). While others have produced excellent overviews of the many clustering approaches available \citep[e.g.,][]{xu2015}, we will briefly introduce some of the more readily available approaches suitable for most time series feature data. The well-established and readily available clustering approaches can, broadly speaking, be categorized as based on (1) centroids, (2) distributions, (3) density, (4) hierarchies, or (5) a combination thereof (see \tblref{tab:clusterApproaches}).

@@

\paragraph{Cluster Performance Evaluation}

@@

incl., finding cluster prototypes:
1. The medoid sequence of the set. 2. The average sequence of the set. 3. The local search prototype

External evaluation measures: These measures require knowledge of the true class labels and are used to compare the cluster assignments to the true class labels. Examples include accuracy, F-measure, adjusted Rand index, mutual information, and normalized mutual information.

Internal evaluation measures: These measures do not require knowledge of the true class labels and are used to evaluate the quality of the clusters based on the characteristics of the data itself. Examples include silhouette coefficient, Calinski-Harabasz index, Davies-Bouldin index, and the Elbow method.
